CARBON CONDENSATION NEAR THE HE SHELL IN SUPERNOVAE

### 2013-07-27 doc/condensation.tex
Modify text.

For a reaction:
\[
1 + 2 \longleftrightarrow 3 + \gamma
\]
we can write the number density change of species 1 as:
\begin{equation}
\frac{{\rm d}n_1}{{\rm d}t} = 
  -k \, n_1 n_2 + \lambda_\gamma n_3
\label{eq:number_density}
\end{equation}
where $k$ is the reaction rate in the unit of $cm^3s^{-1}$ and 
$\lambda_\gamma$ is the photodissociation rate of species 3 with the unit of
$s^{-1}$.

In a expanding or a collapsing system the number densities change with volume. 
We'd like to study a quantity that does not depend on volume directly so that
we use the fraction of each species in number instead of number density:
\[
Y_i \equiv n_i / n_{total}
\]
where $Y_i$ is called the abundance of species i, $n_i$ is the number density
of species i and $n_{total}$ is the number density of the unit particles, here
the number density of total atoms. So $Y_i$ here indicates the number
of species i per atom. And
\[
n_{total} = n_{atom} = n_{baryon} \times Y_{atom} 
\]
where $n_{baryon} = \rho N_A$ is the baryon number density and $Y_{atom}$ is
the atom fraction, i.e. number of atoms per baryon. Take pure $^{12}$C for 
example, $Y_{atom} = 1 / 12$. So Eq.\ref{eq:number_density} can be rewrite as
\begin{equation}
\frac{{\rm d}Y_1}{{\rm d}t} = 
  -k n_{atom} =
  -k \rho N_A Y_{atom} Y_1 Y_2 + \lambda_\gamma Y_3
\label{eq:abundance}
\end{equation}
where $\rho$ is the mass density and $N_A$ is the Avogadro number. Then what
we need to study is how the abundances of each species change with time.

@@

For a single kind of atom, we study the effective condensation rate here.
The condensation here means one atom attaches to an existing multi-atom
particle:
\begin{equation}
P_1 + P_i \longleftrightarrow P_{i+1} + \gamma
\label{eq:reaction}
\end{equation} 
where $P_i$ denotes a particle with $i$ atoms in it.

Let's assume that the condensation rate is proportional to the surface area
(or cross section) of the larger particle, which goes like $n^{2/3}$, i.e.
\[
k \propto A \propto n^{2/3}
\]
where $k$ is the rate, $A$ is the surface area and $n$ is the number of atoms
in the particle.

To make the calculation of a large atom number particle possible, we introduce
the ``bin" after a continuous size increment. In a bin we neglect the details 
of the single particles and focus on the
two ends of the bin. For example if we start the bin from size $a$ and the bin 
size is 10, the particles we care about after size $a$
are just $P_a$, $P_{10a}$, $P_{100a}$, etc. 
And we just calculate the effective reaction rates between these particles.

From graph theory we know that for a series of vertices 1, 2, ..., N, the
effective rate between 1 and N is like:
\[
k_{1,N}^{eff} \Delta t = \frac 
  { k_{1,2}k_{2,3}...k_{N-1,N}\Delta t^{N-1} }
  { 
    k_{2,3}k_{3,4}...k_{N-1,N}\Delta t^{N-2} +
    k_{1,2}k_{3,4}...k_{N-1,N}\Delta t^{N-2} +
    ...
    k_{1,2}...k_{N-2,N-1}\Delta t^{N-2}
  }
\]

\[
k_{1,N}^{eff} = \frac {1}
  { 
    \frac{1}{k_{1,2}} +
    \frac{1}{k_{2,3}} +
    ...
    \frac{1}{k_{N-1,N}}
  }
\]

Back to the problem we are looking at(Eq.\ref{eq:reaction}), we can write
\[
k_{i,i+1} \equiv k_i = k_1 i^{2/3}
\]
If we start from $p_a$, for the bin size of $b$ we can write the effective
reaction as:
\[
P_a + a\times (b-1) P_1 \to P_{a\times b}
\]
And the effective rate is
\[
k^{eff} = \frac {1}
  {
    \frac {1} {k_{a}} +
    \frac {1} {k_{a+1}} +
    ...
    \frac {1} {k_{a\times b}}
  }
\]
\[
\frac {1} {k_{eff}} =
  \sum_{i=a}^{a\times b} \frac {1} {k_i} =
  \frac{1}{k_1} \sum_{i=a}^{a\times b} \frac {1} {i^{2/3}} \approx
  \frac{1}{k_1} \int_a^{a\times b} x^{-2/3} {\rm d} x =
  \frac{1}{k_1} \cdot 3 \cdot [(a\times b)^{1/3} - a^{1/3}]
\]
\[
k^{eff} = \frac{k_1}{3} 
  \frac{1}{ (a\times b)^{1/3} - a^{1/3} }
\]

If we denote the atom number at the beginning of a bin as $N_{begin}$ and
the one at the end as $N_{end}$, then we can write the effective rate as:
\begin{equation}
k^{eff} = \frac{k_1}{3} 
  \frac{1}{ N_{end}^{1/3} - N_{begin}^{1/3} }
\label{eq:rate}
\end{equation}

We estimate $k_1$ as two single atoms collide together with thermal velocity
and the transverse section.
\begin{equation}
k_1 = \: <\! \sigma_1 v\! > \: = \pi R^2 \sqrt{3k_BT/m}
\label{eq:k1}
\end{equation}
where $\sigma_1$ is the cross section for a single atom and $R$ is the radius
of a single atom, $v$ is the thermal velocity which is taken to be the average
value in 3 dimensions, $k_B$ is the Boltzmann constant, $T$ is the temperature
and $m$ is the mass of a single atom.

@@

The abundance change rates of $P_{begin}$, $P_{end}$ and $P_1$ due to 
reaction $P_{begin} + X P_1 \longleftrightarrow P_{end} + \gamma$ are
\[
\frac{{\rm d}Y_{begin}}{{\rm d}t} =
  -k^{eff} Y_{atom} Y_{begin} Y_1 + \lambda_\gamma^{eff} Y_{end}  
\]
\[
\frac{{\rm d}Y_{end}}{{\rm d}t} =
  k^{eff} Y_{atom} Y_{begin} Y_1 - \lambda_\gamma^{eff} Y_{end}  
\]
\[
\frac{{\rm d}Y_1}{{\rm d}t} =
  X( -k^{eff} Y_{atom} Y_{begin} Y_1 + \lambda_\gamma^{eff} Y_{end} )
\]
where $\lambda_\gamma^{eff}$ is the effective photodissociation rate which
will be discussed later and $X$ is the factor to be determined that ensure
the total atom number conserved. Atom conservation requires
\[
N_{begin} \times \frac{{\rm d}Y_{begin}}{{\rm d}t} +
N_{end} \times \frac{{\rm d}Y_{end}}{{\rm d}t} +
\frac{{\rm d}Y_1}{{\rm d}t} = 0.
\] 
So we have
\[
X = N_{end} - N_{begin}
\]

>>>>

For a reaction:
\[
1 + 2 \longleftrightarrow 3 + \gamma
\]
we can write the number density change of species 1 as:
\begin{equation}
\frac{{\rm d}n_1}{{\rm d}t} = 
  -\lambda_{12} \, n_1 n_2 + \lambda_\gamma n_3
\label{eq:number_density}
\end{equation}
where $\lambda_{12}$ is the reaction rate in the unit of $cm^3s^{-1}$ and 
$\lambda_\gamma$ is the photodissociation rate of species 3 with the unit of
$s^{-1}$.

In a expanding or a collapsing system the number densities change with volume. 
We'd like to study a quantity that does not depend on volume directly so that
we use the fraction of each species in number instead of number density:
\[
Y_i \equiv n_i / n_{total}
\]
where $Y_i$ is called the abundance of species i, $n_i$ is the number density
of species i and $n_{total}$ is the number density of the unit particles, here
the number density of total atoms. So $Y_i$ here indicates the number
of species i per atom. And
\[
n_{total} = n_{atom} = n_{baryon} \times Y_{atom} 
\]
where $n_{baryon} = \rho N_A$ is the baryon number density and $Y_{atom}$ is
the atom fraction, i.e. number of atoms per baryon. Take pure $^{12}$C for 
example, $Y_{atom} = 1 / 12$. So Eq.\ref{eq:number_density} can be rewrite as
\begin{equation}
\frac{{\rm d}Y_1}{{\rm d}t} = 
  -\lambda_{12} n_{atom} Y_1 Y_2 + \lambda_\gamma Y_3 =
  -\lambda_{12} \rho N_A Y_{atom} Y_1 Y_2 + \lambda_\gamma Y_3
\label{eq:abundance}
\end{equation}
where $\rho$ is the mass density and $N_A$ is the Avogadro number. Then what
we need to study is how the abundances of each species change with time.

@@

For a single kind of atom, we study the effective condensation rate here.
The condensation here means one atom attaches to an existing multi-atom
particle:
\begin{equation}
P_1 + P_i \to P_{i+1} + \gamma
\label{eq:reaction}
\end{equation} 
where $P_i$ denotes a particle with $i$ atoms in it. Here we are neglecting
the photodissociation for large stable particles.

Let's assume that the condensation rate is proportional to the surface area
(or cross section) of the larger particle, which goes like $N^{2/3}$, i.e.
\[
\lambda \propto A \propto N^{2/3}
\]
or
\[
\lambda_i = \lambda_1 N_i^{2/3}
\]
where $\lambda$ is the rate, $A$ is the surface area and $N$ is the number 
of atoms in the particle.

To make the calculation of particles with a large atom number possible, we 
introduce
the ``bin" after a continuous network. There will a number of particles in a 
bin and we assume \textbf{steady state} within a bin.
The rate out of a bin is like
\[
\lambda_{bin,last}Y_{bin,last} = \lambda^{eff}_{bin}Y_{bin}
\]
\[
Y_{bin} = \sum_{i\in bin} Y_i
\]
\[
\lambda^{eff}_{bin} = \lambda_{bin,last} \left(
  \frac{Y_{bin,last}}{Y_{bin}}
  \right)_{ss}
  = \lambda_{bin,last} \left(
  \frac{1}{\sum_{i\in bin} \frac{Y_i}{Y_{bin,last}} }
  \right)_{ss}
\]
For steady state,
\[
\frac{Y_i}{Y_j} = \frac{\lambda_i}{\lambda_j}
\]
so,
\[
\lambda^{eff}_{bin} = \lambda_{bin,last} \left(
  \frac{1}{\sum_{i\in bin} \frac{\lambda_{bin,last}}{\lambda_i} }
  \right)
  = \frac{1}{\sum_{i\in bin} \frac{1}{\lambda_i} }
  = \lambda_1 \frac{1}{\sum_{i\in bin} N_i^{-2/3}}
\]
If we apply the integral approximation,
\begin{equation}
\lambda^{eff}_{bin}
  \approx \lambda_1 \left( \int_{bin,fist}^{bin,last} x^{-2/3} {\rm d} x
  \right)^{-1}
  = \lambda_1 \frac{1}{N_{bin,last}^{1/3} - N_{bin,first}^{1/3}}
\label{eq:rate}
\end{equation}

We estimate $\lambda_1$ as two single atoms collide together with thermal 
velocity and the transverse section.
\begin{equation}
\lambda_1 = \: <\! \sigma_1 v\! > \: = \pi R^2 \sqrt{3k_BT/m}
\label{eq:k1}
\end{equation}
where $\sigma_1$ is the cross section for a single atom and $R$ is the radius
of a single atom, $v$ is the thermal velocity which is taken to be the average
value in 3 dimensions, $k_B$ is the Boltzmann constant, $T$ is the temperature
and $m$ is the mass of a single atom.

@@

The abundance change rates of $Y_{bin,i}$, $Y_{bin,i+1}$ and $Y_1$
due to 
reaction ${\rm Bin}_i + X P_1 \to {\rm Bin}_{i+1} + \gamma$ are
\[
\frac{{\rm d}Y_{bin,i}}{{\rm d}t} =
  -\lambda_i^{eff} n_{atom} Y_{bin,i} Y_1
\]
\[
\frac{{\rm d}Y_{bin,i+1}}{{\rm d}t} =
  \lambda_i^{eff} n_{atom} Y_{bin,i} Y_1
\]
\[
\frac{{\rm d}Y_1}{{\rm d}t} =
  X( -\lambda_i^{eff} n_{atom} Y_{bin,i} Y_1 )
\]
where $X$ is the factor to be determined that ensure
the total atom number conserved. Atom conservation requires
\[
N_{bin,i} \times \frac{{\rm d}Y_{bin,i}}{{\rm d}t} +
N_{bin,i+1} \times \frac{{\rm d}Y_{bin,i+1}}{{\rm d}t} +
\frac{{\rm d}Y_1}{{\rm d}t} = 0.
\] 
So we have
\[
X = N_{bin,i+1} - N_{bin,i}
\]

Similarly the abundance change rates of $Y_{last}$, $Y_{bin,1}$ and 
$Y_1$ due to reaction $P_{network,last} + X P_1 \to {\rm Bin}_1 + \gamma$ are
\[
\frac{{\rm d}Y_{last}}{{\rm d}t} =
  -\lambda_{last} n_{atom} Y_{last} Y_1
\]
\[
\frac{{\rm d}Y_{bin,1}}{{\rm d}t} =
  \lambda_{last} n_{atom} Y_{last} Y_1
\]
\[
\frac{{\rm d}Y_1}{{\rm d}t} =
  (-\lambda_{last} n_{atom} Y_{last} Y_1) \times
  (N_{bin,1} - N_{last})
\]

\section{Network Matrix}
Let's look at the reaction between the last species in the continuous network
and the first bin, 
the changes of the root-finding functions are (how to call them?)
(denoting $n_{atom}$ as $n_a$):
\[
f_{last} \: +\!\!= \: \lambda_{last} \, n_a Y_{last} Y_1
\]
\[
f_{1} \: +\!\!= \: (\lambda_{last} \, n_a Y_{last} Y_1) 
  (N_{bin,1} - N_{last})
\]
To find the abundance changes in the continuous network, we need to solve the
matrix
\[
AX=B
\]
The matrix element changes due to this reaction are
\[
A_{last,last} \: +\!\!= \: \lambda_{last} \, n_a Y_1
\]
\[
A_{last,1} \: +\!\!= \: \lambda_{last} \, n_a Y_{last}
\]
\[
A_{1,last} \: +\!\!= \: (\lambda_{last} \, n_a Y_1)(N_{bin,1} - N_{last})
\]
\[
A_{1,1} \: +\!\!= \: (\lambda_{last} \, n_a Y_{last})(N_{bin,1} - N_{last})
\]
The right-hand-side vector changes are
\[
B_{last} \: +\!\!= \: -f_{last} = -\lambda_{last} \, n_a Y_{last} Y_1 
\]
\[
B_1 \:+\!\!=\: -f_1 = -(\lambda_{last}\,n_a Y_{last} Y_1)(N_{bin,1}-N_{last})
\]

Now let's look at the reactions between bins. 
${\rm Bin}_i + (N_{bin,i+1}-N_{bin,i})P_1 \to {\rm Bin}_{i+1}$,
where $i$ goes from 1 to $N-1$ with $N$ as the number of bins.
\[
\left( \frac{{\rm d}Y_1}{{\rm d}t}\right) _i =
  ( -\lambda^{eff}_{bin,i} \, n_a Y_{bin,i} Y_1 )(N_{bin,i+1} - N_{bin,i})
\]
\[
(f_{1})_i \: +\!\!= \: 
  (\lambda^{eff}_{bin,i} \, n_a Y_{bin,i} Y_1) (N_{bin,i+1} - N_{bin,i})
\]
\[
(A_{1,1})_i \: +\!\!= \: 
  (\lambda^{eff}_{bin,i} \, n_a Y_{bin,i}) (N_{bin,i+1} - N_{bin,i})
\]
\[
(B_1)_i \:+\!\!=\: -(f_1)_i = 
  -(\lambda^{eff}_{bin,i} \, n_a Y_{bin,i} Y_1) (N_{bin,i+1} - N_{bin,i})
\]
We first solve the matrix equation for current $Y_{bin,i}$, and then compute
the abundance changes in bins using the new abundances in the network 
(as showed below).
Then feed the new bin abundances back to the matrix. Iterate until converge.

\section{Bin Abundances Change}

Now let's look at the abundance changes in bins for each reaction discussed
above. We are going to solve these changes implicitly. For $Y_{bin,1}$,
\[
\frac {Y_{bin,1}(t+\Delta t) - Y_{b1}(t)} {\Delta t} =
  \lambda_{last} \, n_a Y_{last}(t+\Delta t) Y_1(t+\Delta t) -
  \lambda^{eff}_{bin,1} \, n_a Y_{bin,1}(t+\Delta t) Y_1(t+\Delta t)
\]
\[
Y_{bin,1}(t+\Delta t) = \frac
  {\lambda_{last} \, n_a Y_{last}(t+\Delta t) Y_1(t+\Delta t) \Delta t +
    Y_{bin,1}(t)}
  {1 + \lambda^{eff}_{bin,1} \, n_a Y_1(t+\Delta t) \Delta t}
\]
Similarly for $Y_{bin,i}$ as $i$ goes from 2 to $N-1$:
\[
Y_{bin,i}(t+\Delta t) = \frac
  {\lambda^{eff}_{bin,i-1} \, n_a Y_{bin,i-1}(t+\Delta t) Y_1(t+\Delta t) 
    \Delta t +
    Y_{bin,i}(t)}
  {1 + \lambda^{eff}_{bin,i} \, n_a Y_1(t+\Delta t) \Delta t}
\]
For $Y_{bin,N}$ we don't have any destructive terms for now,
\[
Y_{bin,N}(t+\Delta t) = 
  \lambda^{eff}_{bin,N-1} \, n_a Y_{bin,N-1}(t+\Delta t) Y_1(t+\Delta t) 
    \Delta t +
    Y_{bin,N}(t)
\]
